\chapter{English Rationality is eating itself}

Harry Lee Kwan Yew of course mastered English Rationality.


It is now common for people to think that thinking and intellectualism is all about coherence, and that empirical judgements are to be eschewed. This is essentially equivalent to only discussing the consistency of propositions but never their own truth value. That has become a task relegated to the realm of emotions, and therefore private and respectable regardless of content. The English Rationality does this because they want to ground their paradigm out of nothing, so that it can become self evident, a tautology. But the problem of course, is that such activity is impossible. We live, in a contingent world. That contingency might be necessary, but we not living in a necessity. 


Only the English speaking fool could believe this. The English speaker masks his ignorance and stupidity by the sheer rationality of the English language. 



This is what I’ve been saying for the past year. The English speaking world is unfolding, twisted into a knot in its own body by its own impeccable rationality. As far as I see, its demise is inevitable. The hitherto unstoppable expansion of the English empire will now halt, and shall enter a long meandering age of managed decline. What will rise to supplant her? None under heaven I answer. 


It’s as if Britain had turned herself into one of Lewis Carroll’s novels. “We’re all mad here,” proclaimed the nation that gave birth to the rule of law, analytic philosophy, sublime Rationality, and the English language. 


Has the English speaking world walked itself into a knot by its own impeccable and iron-willed analyticitiy, logicalitu, and rationality? I think it has. Not only has the Angloworld been bewitched by its own language’s absolutely stupendous ability to cut through and apart everything so what lies under and beneath is for all the world to see. The subtle English knife ist indeed knife subtle englishly. By the power of analysis, ANALYSIS!- it has now been trained and accumulated with such powers that it can no longer hold its power for proper use but instead finds itself incapable of not analysing everything and anything there is. And in that horrific deranged analysis, it feels an incredible drive, an irresistible drive to conquer all - to make all right. Anger thus sets in. Indignation is therefore become the second tongue to English speakers. 

This cannot be solved except fid a careful loosening is the jnot. But how could that be done? The English language is not a rope, it is a snake. A knot in a rope may be tight but unknot it with force is possible. In the worse case scenario in which there is not in fact an unknotting force but a ever more knotting tightening force, the rope suffers no harm. It stays as it is but only with a tighter knot. But that is not so for a snake. A snake in a knot finding the knot growing tighter and tighter, will also eventually find her bones and skeleton crushed, her various organs squished and gradually squashed, her digestive duct compressed and her blood vessels ploding - if the snake does not already die from this, sooner or later, she snaps. 